Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-14 paper/main.tex
attempts to describe the rnaseq

A class wise evaluation including the calculation of the Genic % still could use a better location...
F1 score is performed on the validation set after every epoch. The best model
from any training run was selected as the model with the highest Genic F1 after either
maximum epochs were reached or Genic F1 stopped improving and training was interrupted. % check / what was patience set to?

@@

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we performed 4 separate animal and plant training runs and 
selected two models each, specifically the model with the best precision and the  % please check
other with the best recall. The model with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by Augustus. An untrained model was setup with "new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with "optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by Augustus were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with Augustus, so the metric Sub Genic F1, which excludes UTR,
was used for all comparisons between Augustus and Helixer.

@@

Finally, the selected, quality, mapped RNAseq samples were quantified to
get the coverage (number of reads matching; i.e. cigar =, M or X),
and spliced coverage (number of reads with gap or splice; i.e. cigar N or D)
for every base pair in the genomes as implemented in the "training\_rnaseq.py"
script.

@@

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized by the 
Figures S1 and S3. We do not, for example, consistently predict very well on the 
algae as well as the non-avian reptiles. While one of the the algae were included 
in our training genomes, the non-avian reptiles were not. Our model also tends to 
show higher uncertainty around transitions of annotation classes (e.g. from UTR to 
CDS) due to the lack of further post-processing of the basepair wise predictions.

AUGUSTUS does outperform us rather consistently on the smallest genomes, but falls 
off much more the larger the genomes get. The difference in prediction quality is 
especially strong for mammals, which tend to have big genomes with very long genes 
as well as the largest plants. Figures S2 and S4 display the comparison to AUGUSTUS 
by phylogenetic position.

Two techniques were used during inference to improve performance and limit model 
bias. The usage of longer input lengths helped especially for genomes that tend to 
have longer genes and was enabled by our model architecture being a relatively 
simple LSTM stack without any fully connected layers on top. We also constructed the 
final predictions out of a overlapping ones, which greatly helped to reduce 
prediction bias in most genomes. To our knowledge, both techniques were not used 
before in a model developed for gene annotation.

>>>>

A class wise evaluation including the calculation of the Genic 
F1 score is performed on the validation set after every epoch. The best model
from any training run was selected as the model with the highest Genic F1 after either
maximum epochs were reached or Genic F1 stopped improving and training was interrupted.

@@

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we performed 4 separate animal and plant training runs and 
selected two models each, specifically the model with the best precision and the  % please check / epoch model / version something
other with the best recall. The model with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by Augustus. An untrained model was setup with "new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with "optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by Augustus were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with Augustus, so the metric Subgenic F1, which excludes UTR,
was used for all comparisons between Augustus and Helixer.

@@

Finally, the selected, quality, mapped RNAseq samples were merged and quantified to
get the coverage (number of reads matching; i.e. cigar =, M or X),
and spliced coverage (number of reads with gap or splice; i.e. cigar N or D)
for every base pair in the genomes as implemented in the "training\_rnaseq.py"
script.

@@

Our models, however, tend to perform less well for the very smallest and largest 
genomes or species that are phylogenetically the furthest away from our training 
genomes. This is the case for both animals and plants and is visualized by the 
Figures S1 and S3. We do not, for example, consistently predict very well on the 
algae as well as the non-avian reptiles. While one of the the algae were included 
in our training genomes it accounted for a tiny proportion of the total training 
data and the non-avian reptiles were not included at all.

AUGUSTUS outperforms us on some of the smallest genomes, but falls off much more 
drastically as the genomes get larger. The difference in prediction quality is 
especially strong for mammals, which tend to have big genomes with very long genes 
as well as the largest plants. Figures S2 and S4 display the comparison to AUGUSTUS 
by phylogenetic position.

Two techniques were used during inference to improve performance and limit model 
bias. The usage of longer input lengths helped especially for genomes that tend to 
have longer genes and was enabled by our model architecture being a relatively 
simple LSTM stack without any fully connected layers on top. We also constructed the 
final predictions out of a overlapping ones, which greatly helped to reduce 
prediction bias in most genomes. To our knowledge, neither technique has been used 
before in a model developed for gene annotation.

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences \ref{fig:examples} % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d). Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation `
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a).

In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (b and c,hSlA2 supp). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (hSlA2, hSlA3, lSlA3). In one case the 
has not made a reasonably is not assigning a sufficiently 
confident probability to any
